\documentclass[11pt]{article}
\usepackage[margin=1in]{geometry}
\usepackage{amsmath}
\usepackage{amssymb}
\usepackage{booktabs}
\usepackage{hyperref}

\title{Lingua Symbolic Prompt Compiler: Optimization Objective}
\author{Lingua Symbolic Prompt Compiler}
\date{}

\begin{document}
\maketitle

\section{Problem}

Let \(M\) be the frozen target model configuration, \(P\) the original prompt
template, and \(X = \{x_i\}_{i=1}^{n}\) the input set. The compiler searches for
a compressed prompt template \(P'_c\) that reduces instruction tokens while
preserving the target model's behavior over \(X\).

\[
y_i = M(P, x_i)
\]

\[
\hat{y}_{i,c} = M(P'_c, x_i)
\]

The outputs \(y_i\) are behavioral references. Candidate quality is measured by
how closely \(\hat{y}_{i,c}\) matches those references under the configured
evaluation metrics.

\section{Template Variables}

Prompt variables are template slots such as \texttt{\{\{input\}\}},
\texttt{\{\{query\}\}}, or \texttt{\{\{document\}\}}. Let \(B(P)\) be the
ordered placeholder sequence extracted from \(P\). Candidate assembly preserves
this sequence:

\[
B(P'_c) = B(P).
\]

The instruction-token objective is measured on the invariant prompt layer,
which is the prompt template with placeholder contents removed.

\section{Token Objective}

Let \(T_{\mathrm{instr}}(P)\) be the configured tokenizer count for the
instruction layer of \(P\). Candidate token ratio is:

\[
\rho_c =
\frac{T_{\mathrm{instr}}(P'_c)}
     {T_{\mathrm{instr}}(P)}.
\]

Instruction-token reduction is:

\[
r_c = 1 - \rho_c.
\]

Rendered prompt cost is also tracked:

\[
\bar{T}_{\mathrm{full}}(P'_c) =
\frac{1}{n}\sum_{i=1}^{n} T(P'_c(x_i)).
\]

\section{Embedding Drift}

Let \(E\) be the configured embedding model. Lingua Symbolic Prompt Compiler uses
raw Euclidean distance over completion embeddings in the configured embedding
model's vector space:

\[
d_{\mathrm{emb}}(c, i) =
\left\| E(\hat{y}_{i,c}) - E(y_i) \right\|_2.
\]

Let \(s_d > 0\) be the semantic-drift normalization scale for the embedding
model and evaluation context. The normalized drift term is:

\[
d_{\mathrm{norm}}(c, i) =
\mathrm{clip}\left(\frac{d_{\mathrm{emb}}(c, i)}{s_d}, 0, 1\right).
\]

Average drift for candidate \(c\) is:

\[
\bar{d}_{\mathrm{emb}}(c) =
\frac{1}{n}\sum_{i=1}^{n} d_{\mathrm{emb}}(c, i).
\]

Average normalized drift is:

\[
\bar{d}_{\mathrm{norm}}(c) =
\frac{1}{n}\sum_{i=1}^{n} d_{\mathrm{norm}}(c, i).
\]

\section{Objective}

The scalar objective is a normalized loss that combines token loss and
generated-output semantic drift:

\[
\tilde{r}_c = \mathrm{clip}(r_c, 0, 1).
\]

\[
\ell_t(c) = 1 - \tilde{r}_c.
\]

\[
\mathcal{L}(c) =
\frac{\lambda_t \ell_t(c) + \lambda_d \bar{d}_{\mathrm{norm}}(c)}
     {\lambda_t + \lambda_d}.
\]

The default equal-weight loss is:

\[
\mathcal{L}(c) =
0.5 \ell_t(c) + 0.5 \bar{d}_{\mathrm{norm}}(c).
\]

The loss is bounded to \([0, 1]\), and lower is better.

\section{Validity Signals}

Structured output validity is tracked separately from the scalar objective.
For a candidate \(c\), let:

\[
\phi_c =
\frac{1}{n}\sum_{i=1}^{n}\mathbf{1}[\text{format failure for }(c,i)]
\]

\[
\tau_c =
\frac{1}{n}\sum_{i=1}^{n}\mathbf{1}[\text{task-field failure for }(c,i)].
\]

The report includes these rates and the associated failure cases so invalid
candidate behavior is visible during review.

Current core objective weights are:

\[
\lambda_t = 0.50,\quad
\lambda_d = 0.50.
\]

\section{Pareto Frontier}

Candidate \(a\) dominates candidate \(b\) when it is at least as good on every
frontier dimension and better on at least one:

\[
r_a \ge r_b,\quad
\bar{d}_{\mathrm{emb}}(a) \le \bar{d}_{\mathrm{emb}}(b).
\]

Validity rates \(\phi_c\) and \(\tau_c\) are reported alongside the frontier.

The returned frontier is the non-dominated set:

\[
\mathcal{F} =
\{c \in C : \nexists c' \in C \text{ such that } c' \succ c\}.
\]

\section{Epoch Flow}

\begin{center}
\begin{tabular}{lll}
\toprule
Layer & Role & Current implementation \\
\midrule
Reference & Build behavioral references & target model calls \\
Bound slots & Preserve template variables & placeholder sequence \\
Chunking & Create rewrite units & para, sent, md, schema, role, window \\
Exploration & Propose compressed chunks & English, DSL, Mandarin, mixed \\
Evaluation & Compare completions & token ratio and output distance \\
Optimization & Advance population & Pareto and mutation \\
\bottomrule
\end{tabular}
\end{center}

\section{Dataset Split}

For \(n > 3\), the reference dataset is split deterministically:

\[
X_{\mathrm{train}} = X_{1:\lfloor 0.6n \rfloor},
\quad
X_{\mathrm{dev}} = X_{\lfloor 0.6n \rfloor:\lfloor 0.8n \rfloor},
\quad
X_{\mathrm{holdout}} = X_{\lfloor 0.8n \rfloor:n}.
\]

The optimizer searches on train subsets, selects finalists on dev, and reports
holdout behavior for final frontier candidates.

\end{document}
